Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with \pkg{ACSNMineR} and \pkg{RNaviCell} packages

### 2016-08-30 jss/Deveau.tex
[skip ci] text update, multiple paragraphs

%% an abstract and keywords
\abstract{
Biological pathways or modules represent sets of interactions or functional relationships occurring at the molecular level in living cells. A large body of knowledge on pathways is organized in public databases such as the KEGG, Reactome, or in more specialized repositories, the Atlas of Cancer Signaling Network (ACSN) being an example. All these open biological databases facilitate analyses, improving our understanding of cellular systems. We hereby describe \CRANpkg{ACSNMineR} for calculation of enrichment or depletion of lists of genes of interest in biological pathways. \pkg{ACSNMineR} integrates ACSN molecular pathways, but can use any molecular pathway encoded as a GMT file, for instance sets of genes available in the Molecular Signatures Database (MSigDB). We also present \CRANpkg{RNaviCell}, that can be used in conjunction with \pkg{ACSNMineR} to visualize different data types on web-based, interactive ACSN maps. We illustrate the functionalities of the two packages with biological data taken from large-scale cancer datasets.

@@

Only the gene list is mandatory to call the \code{enrichment} function, all the
other arguments have default values.  The \code{maps} argument can either be a
dataframe imported from a gmt file with the \code{format\_from\_gmt} function or
a list of dataframes generated by the same procedure. By default, the function
uses the ACSN maps previously described.  The correction for multiple testing
is set by default to use the method of Benjamini \& Hochberg, but can be
changed to any of the usual correction methods (Bonferroni, Holm, Hochberg,
Holm, or Benjamini \& Yekutieli \citep{Benjamini2003FDR}), or even disabled .
The minimal module size represents the smallest size value of a module that
will be used to compute enrichment or depletion. This is meant to remove
results of low significance for module of small size.  The universe in which
the computation is made by default is defined by all the gene symbols contained
in the maps. All the genes that were given as input and that are not present on
the maps will be discarded. To keep all genes, the user can change the universe
to \code{HUGO}, and in that case, the complete list of HUGO gene symbols will
be used as the reference ($>$ 39,000 genes). The threshold corresponds to the
maximal value of the corrected p-value (unless the user chose not to correct
for multiple testing) that will be displayed in the result table.

@@

In the simple example detailed below, we create a NaviCell session, then load
an expression data set from a local (tab-delimited) file. The data represent
gene expression measured in a prostate cancer cell line resistant to hormonal
treatment (agressive), and is taken from the Cell Line Encyclopedia project
\citep{barretina2012cancer}. We visualize the data values on the Cell Cycle map
(the default map), using heat maps. With this visualization mode, gene
expression values are represented as a color gradient (green to red) in
squares positioned next to the entities where the gene has been mapped (Figure~\ref{fig:du145}).

@@

The Molecular Signatures Database (MSigDB) is one of the most widely used
repository of well-annotated  gene sets representing the universe of biological
processes \citep{liberzon2011molecular}. We downloaded the canonical pathways
set, counting more than 1,300 gene sets representing canonical pathways compiled
by domain experts. The dataset is encoded with the 'gmt' format, and can be
imported within ACSNMineR with the \code{format\_from\_gmt} function. We
calculate the enrichment for the breast cancer differentially expressed gene
list, simply specifying the MSigDB data we just imported as the \code{maps}
option. Table \ref{tab:table_msigdb} is displaying the pathways having a
corrected p-value $< 0.05$. The prefix is indicating the database source, so we
see that we have pathways from the KEGG, Reactome and PID databases. Consistent
with our previous results, most of the enriched pathways are related to the cell
cycle regulation.

@@

At last, we visualize the mean expression values for ER negative samples for all
genes differentially expressed on the ACSN master (global) map using \code{ACSNMineR}
commands to create heatmaps.

@@

We also describe in this work the \pkg{RNaviCell} package, a R package
convenient to use with \pkg{ACSNMineR}. This package is dedicated to create
web-based and interactive data visualization on ACSN maps. Users can use this
tools to represent genes of interest that have been shown to be related to the
maps by calculating enrichment with the  \pkg{ACSNMineR}.  Creating maps with
the graphical user interface of the ACSN website can be a tedious task if the
user has multiple samples or gene lists, and wants to compare their
representations on ACSN maps. The \pkg{RNaviCell} package can be used to
automate the process of creating the graphical representations automatically.

>>>>

%% an abstract and keywords
\abstract{
Biological pathways or modules represent sets of interactions or functional
relationships occurring at the molecular level in living cells. A large body of
knowledge on pathways is organized in public databases such as the KEGG,
Reactome, or in more specialized repositories, the Atlas of Cancer Signaling
Network (ACSN) being an example. All these open biological databases facilitate
analyses, improving our understanding of cellular systems. We hereby describe
\CRANpkg{ACSNMineR} for calculation of enrichment or depletion of lists of
genes of interest in biological pathways. \pkg{ACSNMineR} integrates ACSN
molecular pathways gene sets, but can use any gene set encoded as a GMT file,
for instance sets of genes available in the Molecular Signatures Database
(MSigDB). We also present \CRANpkg{RNaviCell}, that can be used in conjunction
with \pkg{ACSNMineR} to visualize different data types on web-based,
interactive ACSN maps. We illustrate the functionalities of the two packages
with biological data taken from large-scale cancer datasets.

@@

Only the gene list is mandatory to call the \code{enrichment} function, all the
other arguments have default values.  The \code{maps} argument can either be a
dataframe imported from a GMT file with the \code{format\_from\_gmt} function
or a list of dataframes generated by the same procedure. The GMT format
corresponds to the Broad Institute's Gene Matrix Transposed file format, a
convenient and easy way to encode named sets of genes of interest in
tab-delimited text files (it is not a graph or network format). By default, the
function \code{enrichment} uses the ACSN maps previously described.  The
correction for multiple testing is set by default to use the method of
Benjamini \& Hochberg, but can be changed to any of the usual correction
methods (Bonferroni, Holm, Hochberg, Holm, or Benjamini \& Yekutieli
\citep{Benjamini2003FDR}), or even disabled .  The minimal module size
represents the smallest size value of a module that will be used to compute
enrichment or depletion. This is meant to remove results of low significance
for module of small size.  The universe in which the computation is made by
default is defined by all the gene symbols contained in the maps. All the genes
that were given as input and that are not present on the maps will be
discarded. To keep all genes, the user can change the universe to \code{HUGO},
and in that case, the complete list of HUGO gene symbols will be used as the
reference ($>$ 39,000 genes). The threshold corresponds to the maximal value of
the corrected p-value (unless the user chose not to correct for multiple
testing) that will be displayed in the result table.

@@

In the simple example detailed below, we create a NaviCell session, then load
an expression data set from a local (tab-delimited) file. The data represent
gene expression measured in a prostate cancer cell line resistant to hormonal
treatment (agressive), and is taken from the Cell Line Encyclopedia project
\citep{barretina2012cancer}. We visualize the data values on the Cell Cycle map
(the default map), using heat maps. With this visualization mode, gene
expression values are represented as a color gradient (green to red) in squares
positioned next to the entities where the gene has been mapped
(Figure~\ref{fig:du145}). Note that the map is displayed in a browser and is
\emph{interactive}, i.e users can zoom in to display more information and for
example look in what reactions are involved the genes selected to be displayed,
and lots of other informations (see \cite{bonnet2015navicell} and
\cite{kuperstein2015atlas} for more details).

@@

The Molecular Signatures Database (MSigDB) is one of the most widely used
repository of well-annotated  gene sets representing the universe of biological
processes \citep{liberzon2011molecular}. We downloaded the canonical pathways
set, counting more than 1,300 gene sets representing canonical pathways compiled
by domain experts. The dataset is encoded with the GMT format, and can be
imported within ACSNMineR with the \code{format\_from\_gmt} function. We
calculate the enrichment for the breast cancer differentially expressed gene
list, simply specifying the MSigDB data we just imported as the \code{maps}
option. Table \ref{tab:table_msigdb} is displaying the pathways having a
corrected p-value $< 0.05$. The prefix is indicating the database source, so we
see that we have pathways from the KEGG, Reactome and PID databases. Consistent
with our previous results, most of the enriched pathways are related to the cell
cycle regulation.

@@

At last, we visualize the mean expression values for ER negative samples for all
genes differentially expressed on the ACSN master (global) map using \code{RNaviCell}
commands to create heatmaps.

@@

We also describe in this work the \pkg{RNaviCell} package, a R package
convenient to use with \pkg{ACSNMineR}. This package is dedicated to create
web-based and interactive data visualization on ACSN maps. Users can use this
tools to represent genes of interest that have been shown to be related to the
maps by calculating enrichment with \pkg{ACSNMineR}.  Creating maps with the
graphical user interface of the ACSN website can be a tedious task if the user
has multiple samples or gene lists, and wants to compare their representations
on ACSN maps. The \pkg{RNaviCell} package can be used to automate the process
of creating the graphical representations automatically.  The maps are
displayed in a browser and are interactive, with the possibility for the user
to zoom in and out, search for genes or molecular species, and see the details
of the molecular reactions (what partners are involved, what is the state of a
given species, etc.). For more details on how to use the interface and the
different possibilities, see \cite{kuperstein2013navicell},
\cite{bonnet2015navicell} and \cite{kuperstein2015atlas}.
